---
title: "Get started with mfbdfm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with mfbdfm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mfbdfm)
```

## Introduction and the data

`mfbdfm` combines indicators observed at different frequencies — weekly,
monthly, quarterly — into a single high-frequency activity factor, and
can anchor that factor to a low-frequency target series so it reads
directly as the target's growth rate. The package ships a flagship
application, the **Weekly Activity Index (WAI)**, a weekly GDP indicator
for Switzerland, which this vignette uses as the worked example
throughout.

This vignette is purely applied: it walks the workflow from data to
nowcast and does not state the model. For the measurement and state
equations, the mixed-frequency aggregation weights, identification and
stochastic volatility, see `vignette("methodology", package = "mfbdfm")`.

Two datasets ship with the package. `data_ch_dataset` is the full
curated indicator set; `data_ch_dataset_test` is the variant used here,
and is the one that **also contains the GDP target series** (the full
dataset deliberately does not — real analyses inject GDP at runtime from
the real-time vintage database, see "Mixed frequency" below). Both are
lists split into `flows` and `stocks`:

```{r data}
data(data_ch_dataset_test)

lengths(data_ch_dataset_test)

target <- "ch.seco.gdp.real.gdp.ssa"
```

The split matters: whether a series is a *flow* (a quantity accumulated
over the period — GDP, transactions, traffic counts) or a *stock* (a
level measured at or averaged within the period — prices, indices,
survey balances) selects its temporal aggregation weights, so
misclassifying a series changes the results.

Rather than assembling the two lists by hand, pass the series and a
`meta` table to `mfbdfm_data()`. It accepts a long or wide data frame,
an `mts`, or a list of `ts` — whatever your spreadsheet or database
query produced:

```{r data-input}
series <- c(data_ch_dataset_test$flows[c(target, "SWISSMI", "traffic_PW")],
            data_ch_dataset_test$stocks[1:2])
series <- lapply(series, stats::window, start = 2018)

meta <- data.frame(series = names(series),
                   type = c("flow", "flow", "flow", "stock", "stock"))

d <- mfbdfm_data(series, meta, target = target)
d
```

The reason to prefer this is the printout. When a series' type is
expressed by *which argument it was passed in*, there is nothing to
inspect; printing the object shows the resolved classification and the
frequencies actually in use **before** committing to a run that can take
minutes to hours.

## Estimation and exploration

`ind_dfm()` takes the `mfbdfm_data` object as its first argument. Real
analyses use the full indicator set and a much longer chain — the
accompanying paper uses `length_sample = 5000` after a burn-in of 1000 —
whereas here a five-series subset and a short chain keep this vignette
quick to build:

```{r fit, results = "hide", message = FALSE}
set.seed(1)
fit <- ind_dfm(d, length_sample = 200, burn_in = 50)
```

(`flows`/`stocks` still work as a pair of lists, if you already have
them in that shape; the `mfbdfm_data` argument is an additional entry
point, not a replacement.)

Printing the fit gives the shape of the model and the most recent
nowcasts:

```{r print}
fit
```

`summary()` adds the posterior means of the loadings and the measurement
error standard deviations, plus the residual fit on the observed
entries. The target's loading is fixed at 1 by construction — that
restriction is what makes the factor a GDP growth rate rather than a
generic index — and its measurement error is shrunk toward zero:

```{r summary}
summary(fit)
```

`plot()` shows the factor with a 95% posterior credible band:

```{r plot, fig.width = 6, fig.height = 4}
plot(fit)
```

The standard extractors all work. `coef()` returns the loadings,
`as.data.frame()` gives a tidy frame of the factor and its band, and
`fitted()`/`residuals()` return the augmented dataset and its residuals
with missing observations masked:

```{r extractors}
round(coef(fit), 3)

head(as.data.frame(fit), 3)
```

## Nowcasts

`fit$factor` is the annualized weekly growth rate implied by the model —
the WAI itself — and `fit$nowcast` is the same information at the
target's own (here quarterly) frequency, back in the target's original
units:

```{r nowcast}
window(fit$nowcast, start = 2024)
```

`fit$nowcast_var` carries the posterior variance of each nowcast, for
credible intervals. Note that it is essentially zero in quarters where
the target is actually observed — the identifying prior shrinks the
target's measurement error toward zero, so the model has nothing to be
uncertain about there. It is informative for the periods the model has
to fill in, which is where a nowcast interval is wanted anyway.

`fit$index` is the cumulated activity level, rebased to 100 at the start
of the estimation window — the level counterpart of the growth-rate
factor:

```{r index}
round(window(fit$index, start = c(2025, 40)), 2)
```

## Mixed frequency

Nothing above had to be said about frequencies: the five series enter at
4, 12 and 48 observations per year and the model works on the highest of
them, treating everything unobserved — including the eleven weeks in
which quarterly GDP says nothing — as a latent state. That is the whole
point of the model, and it is why `fit$factor` is weekly even though the
target is quarterly:

```{r frequencies}
c(factor = frequency(fit$factor), nowcast = frequency(fit$nowcast))
```

The mixed-frequency case that needs more care is a *real-time* one,
where the question is not only which frequency a series has but what was
actually published at a given date. The real-time GDP vintage database
ships with the package and needs no configuration:

```{r vintages}
vintages <- get_real_time_gdp_vintages("quarterly")
dim(vintages)
```

`cut_data_real_time()` truncates a dataset to what would have been
observable at a given date, substituting the appropriate GDP vintage for
the target series — the building block for the paper's real-time
out-of-sample evaluation:

```{r cut-real-time}
dat_rt <- cut_data_real_time(data_ch_dataset_test, current_date = 2024.5,
                             GDP_gr_vintages = vintages)
tail(time(dat_rt$flows[[target]]))
```

## Stochastic volatility and serial correlation

Two model features are switchable, and both default to `TRUE` because
both are wanted on real data.

**Stochastic volatility** lets the variance of the common shocks move
over time, so an abrupt swing such as the COVID-19 pandemic registers as
a genuinely more volatile period instead of being smoothed away. With
`stochastic_volatility = FALSE` a single constant variance is estimated
instead:

```{r fit-const, results = "hide", message = FALSE}
set.seed(1)
fit_const <- ind_dfm(d, length_sample = 200, burn_in = 50,
                     stochastic_volatility = FALSE)
```

`fit$pars$h` is the posterior mean of the factor's log innovation
standard deviation, so `exp(2 * h)` is the innovation variance. Under
stochastic volatility it varies over the sample; with the option off it
is flat by construction:

```{r sv-compare}
signif(range(exp(2 * fit$pars$h)), 3)
signif(range(exp(2 * fit_const$pars$h)), 3)
```

Note that switching it off does **not** fix the variance at a value in
`ind_dfm()` — the loading restriction already pins the factor's scale,
so the variance is the free parameter the data determines and it is
still estimated, just held constant over time. In `fcast_dfm()`, whose
loadings are unrestricted, the same option *does* fix it at one. The
methodology vignette's "Identification" section explains why.

**Serial correlation** in the measurement errors is removed by
quasi-differencing, with one autocorrelation per series drawn from the
data. `fit$pars$rho` reports them; a series with a large `rho` is one
whose own persistent noise would otherwise have been attributed to the
common factor:

```{r rho}
round(stats::setNames(as.numeric(fit$pars$rho), fit$inventory$key), 3)
```

`serial_correlation = FALSE` holds all of them at (effectively) zero.

## Beyond this vignette

For several common factors rather than one target-anchored factor, use
`fcast_dfm()`. It implements the multi-factor model of Eckert et al.
(2025) and takes the same data, but identifies the factors by a post-hoc
rotation rather than by anchoring, so its `target` argument only selects
which nowcast is surfaced. `fcast_dfm(q = 1)` is **not** equivalent to
`ind_dfm()`; see `?fcast_dfm` and the methodology vignette.

The functions used in the accompanying analysis pipeline are exported
but need either the full (non-shipped) indicator set or previously saved
model fits, so they aren't run here:

```{r pipeline, eval = FALSE}
# Estimate the full WAI and optionally save the fit:
fit_full <- run_wai_adj(
  flows = dat$flows, stocks = dat$stocks, target = target,
  date = 2024.5, dataset_used = "full_RT", output_dir = "fits/updated"
)

# AR(1) benchmark, and the Diebold-Mariano test used to compare them:
fit_ar <- run_ar(flows = dat$flows, stocks = dat$stocks, target = target,
                 date = 2024.5, dataset_used = "full_RT")
dm_test_modified(errors_wai, errors_ar)
```

For the in-sample and out-of-sample evaluation tables and plots shown in
the paper (correlation heatmaps, relative RMSE/MAE tables against the
SECO-WEA, F-CURVE, SECO-SEC, SNB-BCI and KOF-BARO benchmarks), see the
`analysis/5_plots/` scripts in the package's source repository, which
call `get_combined_cor_table()`, `get_insample_fit_table()`,
`create_rel_error_tables()` and related functions documented under
`?get_combined_cor_table`.

## Conclusion

The workflow is short: describe the series with `mfbdfm_data()`, fit
with `ind_dfm()` (or `fcast_dfm()`), and read the results off the fit
with `summary()`, `plot()` and the standard extractors. What the package
adds over a generic factor model is the mixed-frequency handling, which
needs no imputation step from you, and the target anchoring, which makes
the extracted factor a growth rate you can quote rather than an index
you have to rescale.

Two things to carry into a real run: use a chain far longer than the one
above (the paper's is `length_sample = 5000`, `burn_in = 1000`, and takes
minutes to hours), and check `mfbdfm_data()`'s printed flow/stock split
before starting it.

For the model itself, see
`vignette("methodology", package = "mfbdfm")`.

## References

### Methodology / WAI background

- Kronenberg, P. (2026) — *A high-frequency GDP indicator for
  Switzerland*, Swiss Journal of Economics and Statistics, 162:10.
  <https://doi.org/10.1186/s41937-026-00157-w>. The primary methodology
  and application paper for this package: derives the WAI as a single,
  GDP-identified factor, with full in-sample and real-time
  out-of-sample evaluation against the benchmarks listed below.
- Eckert, F., Kronenberg, P., Mikosch, H., & Neuwirth, S. (2025) —
  *Tracking economic activity with alternative high-frequency data*,
  Journal of Applied Econometrics, 40(3), 270-290.
  <https://doi.org/10.1002/jae.3104>. The multi-factor Bayesian
  mixed-frequency dynamic factor model implemented by `fcast_dfm()`.
- Eckert, Kronenberg, Mikosch, Neuwirth — *Weekly Activity Index (WWA)*,
  SECO technical note and press material, 2021. Available from SECO
  (seco.admin.ch).
- SECO — *Die Wöchentliche Wirtschaftsaktivität (WWA)*, Diskussionspapier.
- SECO — *Konjunkturtendenzen*, Exkurs on the WWA, issue 2020/4.
- SECO — *BIP-Flash Machbarkeitsstudie*, May 2024.
- SECO — *Einfliessende Indikatoren* (WWA input indicator list).
- SECO — *Methodik* note.

The references for the model's derivation — Chan & Jeliazkov (2009),
Chib & Greenberg (1994), Mariano & Murasawa (2003), Bai & Wang (2015),
Kim, Shephard & Chib (1998), Primiceri (2005), Indergand & Leist (2014)
— are listed in `vignette("methodology", package = "mfbdfm")`.

### Swiss business-cycle indicator benchmarks

The in-sample and out-of-sample evaluations compare the WAI against
these existing Swiss business-cycle indicators:

- Glocker, C. and Kaniovski, S. (2018) — *Evaluation of Swiss Business
  Cycle Indicators*, WIFO.
- Glocker, C. and Wegmüller, P. (2019) — *30 Indikatoren auf einen
  Schlag*, Die Volkswirtschaft.
- Wegmüller, P. and Glocker, C. (2024) — *Capturing Swiss Economic
  Confidence*.
- Abberger, K. et al. (2014) — *The KOF Economic Barometer*, KOF, ETH
  Zurich.
- Abberger, K. et al. (2018) — *Using rule-based updating procedures to
  improve the performance*.
- Indergand, R. and Leist, S. (2014) — *A Real-Time Data Set for
  Switzerland*.
- Siliverstovs, B. (2011) — *The Real-Time Predictive Content*.

### Official statistics documentation

- FSO — national accounts documentation.
- SNB — Quartalsbulletin 2018/1.
